Point estimates comparison for the simplified Naomi model

### 2025-05-26 src/docs_paper-jtb/paper.Rmd
Cafe revisions!

Models are specified in `TMB` using a C++ template file which evaluates $\log p(\mathbf{y}, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$.
Other software packages have been developed which also use `TMB` C++ templates.
The `tmbstan` R package [@monnahan2018no] allows running the Hamiltonian Monte Carlo (HMC) algorithm via `Stan`.
The `aghq` R package [@stringer2021implementing] allows use of AGHQ, and AGHQ over the marginal Laplace approximation, via the `mvQuad` R package [@weiser2016mvquad].
The `glmmTMB` R package [@brooks2017glmmTMB] allows specification of common GLMM models via a formula interface.
It is also possible to extract the `TMB` objective function used by `glmmTMB`, which may then be passed into `aghq` or `tmbstan`.
This discussion highlights the utility of `TMB` templates, emphasising that our approach, built on `TMB`, is both general and broadly applicable.

@@

Comparison to further Bayesian inference methods could prove beneficial.
Four possibilities stand out as being particularly valuable.
First, there exist other quadrature rules for moderate dimension, such as the Box-Wilson central composite design [@box1992experimental].
It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
Second, rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
This method can be run by using `tmbstan` and setting `laplace = TRUE`.
When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
Third, NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
It may be difficult to implement such algorithms using `TMB`.
Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language [@de2017programming], however requiring rewriting the Naomi model log-posterior outside of `TMB` would be a downside.
Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.

@@

PCA-AGHQ is a sensible approach to allocating more computational effort to dimensions which contribute more to the integral in question.
Its application to Naomi surfaced several ideas for improvement and challenges to be overcome.
First, the amount of variation explained in the Hessian matrix may not be of direct interest.
For the Naomi model, interest is in the effect of including each dimension on the relevant model outputs.
As such, using alternative measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] or Sobol indices, could be preferable.
Second, use of PCA is challenging when the dimensions have different scales.
For the Naomi model, logit-scale hyperparameters were systematically favoured over those on the log-scale.
Third, when the quadrature rule is used within an INLA algorithm, it is more important to allocate quadrature nodes to those hyperparameter marginals which are non-Gaussian.
This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficient.
The difficulty is knowing in advance which marginals will be non-Gaussian.
This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
Another approach would be to measure the fit of marginal samples from a cheap approximation, like EB.
The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.

>>>>

Models are specified in `TMB` using a C++ template file which evaluates $\log p(\mathbf{y}, \mathbf{x}, \boldsymbol{\mathbf{\theta}})$.
Other software packages have been developed which also use `TMB` C++ templates.
The `tmbstan` R package [@monnahan2018no] allows running the Hamiltonian Monte Carlo (HMC) algorithm via Stan [@rstan].
The `aghq` R package [@stringer2021implementing] allows use of AGHQ, and AGHQ over the marginal Laplace approximation, via the `mvQuad` R package [@weiser2016mvquad].
The `glmmTMB` R package [@brooks2017glmmTMB] allows specification of common GLMM models via a formula interface.
It is also possible to extract the `TMB` objective function used by `glmmTMB`, which may then be passed into `aghq` or `tmbstan`.
This discussion highlights the utility of `TMB` templates, emphasising that our approach, built on `TMB`, is both general and broadly applicable.

@@

Comparison to further Bayesian inference methods could prove beneficial.
Four possibilities stand out as being particularly valuable.
First, there exist other quadrature rules for moderate dimension, such as the Box-Wilson central composite design [@box1992experimental].
It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
Second, rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
This method can be run by using `tmbstan` and setting `laplace = TRUE`.
When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
Third, NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
It may be difficult to implement such algorithms using `TMB`.
Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a subtantial downside.
Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.

@@

PCA-AGHQ is a sensible approach to allocating more computational effort to dimensions which contribute more to the integral in question.
Its application to Naomi surfaced several ideas for improvement and challenges to be overcome.
First, the amount of variation explained in the Hessian matrix may not be of direct interest.
For the Naomi model, interest is in the effect of including each dimension on the relevant model outputs.
As such, using alternative measures of importance from sensitivity analysis, such as Shapley values [@shapley1953value] or Sobol indices, could be preferable.
Second, use of PCA is challenging when the dimensions have different scales.
For the Naomi model, logit-scale hyperparameters were systematically favoured over those on the log-scale.
Third, when the quadrature rule is used within an INLA algorithm, it is more important to allocate quadrature nodes to those hyperparameter marginals which are non-Gaussian.
This is because the Laplace approximation is exact when the integrand is Gaussian, so a single quadrature node is sufficient.
The difficulty is knowing in advance which marginals will be non-Gaussian.
This could be done if there were a cheap way to obtain posterior means, which could then be compared to posterior modes obtained using optimisation.
Another approach would be to measure the fit of marginal samples from a cheap approximation, like EB.
The measures of fit would have to be for marginals, ruling out approaches like PSIS [@yao2018yes] which operate on joint distributions.
Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g. achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].

### 2025-05-26 src/docs_paper-jtb/revisions.Rmd
Cafe revisions!

@@

>>>>

> On `tmbstan` as compared with e.g. `rstan`

`tmbstan` passes the objective function and its gradient (calculated in `TMB`) directly to `rstan` via the [`op_matrix_vari`](https://mc-stan.org/math/classstan_1_1math_1_1op__matrix__vari.html).
As such, the NUTS algorithm used should be the same.
Nonetheless, it could be that there are speed differences between function evaluations in `TMB` and Stan.
We have added a sentence to the discussion noting this possible comparison: "Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder]."

> Questions about the HIV model

Jeff broadly agrees with the reviewer.
(This is an area of current work on Naomi extension.)

> Selecting dimensions in proportion to variance in some outcome explained

It is possible to obtain derivative information about a quantity in a `TMB` model using `ADREPORT`.

> Bayesian optimisation as a competitor in the space

The goal of Bayesian optimisation (BO) is to find the maximum of a (black box) function.
For this reason, we do not see BO as a competitor to algorithms (like ours) intended to perform Bayesian inference.
(One could imagine using BO to optimise the hyperparameters, however it would be preferable to use gradients [as they are available].)

@@

> On Figure 5

Thank you for the suggestion.
We have adjusted the `alpha` value from 0.5 to 0.1 and added a caption to the figure explaining the possible overlap to make the figure clearer:

```{r}
knitr::include_graphics("revisions/fig5_edit.png")
```

> On Figure 6

Yes that is correct, it's the NUTS estimate minus the GEB or GPCA-AGHQ estimate.
We have clarified the $y$-axis here (and where relevant throughout the manuscript) to read "Absolute difference (NUTS - method)".

> On Figure 7

We agree that there are too many colours here for this figure to be clear.
Thank you for the suggestion to split the plot into multiple panels.
We have done this, choosing to group the parameters by the component of the model they are most associated with.
(Remains to do this!)

> ...the method is tunable with how many PCA components are selected and what value of $k$ to use...

We agree that it would be of interest to explore the effect of choosing a different number of PCA components ($s$) and number of quadrature points per dimension ($k$).
One could also consider a varying number of quadrature points.
For example "gradually lowering $k$ over the principal components" presumably would be the most efficient approach.

> ...a useful intermediate scheme to me seems to be to drop the corners of the grid...

We agree that this could be a useful approach.
Indeed Section 6.1.2. reads "Finally, it is likely possible to achieve better performance by pruning and prerotation of the quadrature grid, as discussed by @jackel2005note.".
We believe the suggested approach is what we refer to here (very briefly) as "pruning".
To clarify this point, we have rewritten this sentence as: "Finally, it is likely possible to achieve better performance by pruning (dropping points where the weight is below some threshold) or prerotation (before adaption, in order to e.g. achieve better coverage along the principal axis) of the quadrature grid [@jackel2005note].".

### 2025-05-26 src/naomi-simple_point-estimates/point-estimates.Rmd
Cafe revisions!

df_plot_latent %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_point(shape = 1, alpha = 0.4) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "Estimate", y = "Absolute difference to NUTS") +
  theme_minimal()

@@

mean_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior mean") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.4, 0.4)) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior mean"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference to NUTS") +
  theme_minimal()

@@

sd_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior SD") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.6, 0.6)) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior SD"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference to NUTS") +
  theme_minimal()

@@

df_plot_output %>%
  ggplot() +
  geom_point(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

mean_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior mean") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior mean"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
  theme_minimal()

@@

sd_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior SD") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior SD"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
  theme_minimal()

@@

plot_outputs <- function(.indicator, .par, col) {
  df_plot_output %>%
  filter(indicator == .indicator, par == .par) %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c(col, "white", "white")) +
  facet_grid(indicator + par ~ method, labeller = labeller(indicator = function(x) {rep("", length(x))})) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4)) +
  geom_text(
    data = filter(df_metrics_output_split, indicator == .indicator, par == .par),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference to NUTS", col = "Output") +
  theme_minimal() +
  theme(legend.position = "none", panel.spacing.x = unit(6, "mm"))
}

>>>>

df_plot_latent %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_point(shape = 1, alpha = 0.4) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)") +
  theme_minimal()

@@

mean_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior mean") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.4, 0.4)) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior mean"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)") +
  theme_minimal()

@@

sd_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior SD") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.6, 0.6)) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior SD"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)") +
  theme_minimal()

@@

df_plot_output %>%
  ggplot() +
  geom_point(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

mean_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior mean") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior mean"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
  theme_minimal()

@@

sd_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior SD") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c("#56B4E9","#009E73", "#CC79A7")) +
  facet_grid(indicator ~ method) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior SD"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
  theme_minimal()

@@

plot_outputs <- function(.indicator, .par, col) {
  df_plot_output %>%
  filter(indicator == .indicator, par == .par) %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = c(col, "white", "white")) +
  facet_grid(indicator + par ~ method, labeller = labeller(indicator = function(x) {rep("", length(x))})) +
  geom_abline(intercept = 0, slope = 0, col = "#E69F00", linetype = "dashed") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4)) +
  geom_text(
    data = filter(df_metrics_output_split, indicator == .indicator, par == .par),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "Estimate", y = "Absolute difference (NUTS - method)", col = "Output") +
  theme_minimal() +
  theme(legend.position = "none", panel.spacing.x = unit(6, "mm"))
}
